---
title: "Los Angeles Open Source Geo Hub Data"
subtitle: "Basic Leaflet Work Sample usin APIs" 
author: "Allyson Fierro"
date: "2018"
output:
  prettydoc::html_pretty: default
  html_document: default
---

```{r echo=FALSE}
#Loading libraries
library(rgdal)
library(knitr)
library(leaflet)
library(leaflet.extras)
library(dplyr)
library(readxl)
library(stringr)
```

```{r echo=FALSE}
#Open source Los Angeles, CA GeoHub data 

#Historic earthquakes (points)
quakes.point <- readOGR("https://opendata.arcgis.com/datasets/373a48047cd4473fbb0d02a878c9d13f_62.geojson")

#High injury network (lines)
hwinjury.lines <- readOGR("https://opendata.arcgis.com/datasets/4ba1b8fa8d8946348b29261045298a88_0.geojson")

#Demographic data (polygons)
demog.poly <- readOGR("https://opendata.arcgis.com/datasets/2f11013d1302419aac243d6e1c56568f_0.geojson")
```

$\textbf{Historic earthquakes in LA County}$

This map shows clusters of historic earthquakes in Los Angeles County. Zoom in to view smaller clusters or the location of a specific earthquake. You can also select which base map you would like to view on the top right of the map. 
```{r echo=FALSE}
leaflet(quakes.point) %>%
  addProviderTiles("Esri.WorldImagery", group = "Base") %>%
  addProviderTiles("Esri.WorldTerrain", group = "Terrain") %>%
  addProviderTiles("CartoDB.DarkMatter", group = "DarkMatter") %>%
  addLayersControl(
    baseGroups = c("Base", "Terrain", "Dark Matter"),
    options = layersControlOptions(collapsed = FALSE)) %>%
  addMarkers(
    clusterOptions = markerClusterOptions()
)
```

$\textbf{High Injury Network}$

The injury network represent a small amount of streets, but many serious injuries including death. The network is to provide a guide to help prioritize efforts for achieving zero traffic deaths. 

Zoom in to view street names and select which base map you would like to view on the top right of the map. 
```{r echo=FALSE}
leaflet(data = hwinjury.lines) %>%
  addProviderTiles("Esri.WorldImagery", group = "Base") %>%
  addProviderTiles("Esri.WorldTerrain", group = "Terrain") %>%
  addProviderTiles("CartoDB.DarkMatter", group = "DarkMatter") %>%
  addLayersControl(
    baseGroups = c("Base", "Terrain", "Dark Matter"),
    options = layersControlOptions(collapsed = FALSE)) %>%
  addPolylines(color = "red")
```

$\textbf{Hispanic/Latino Demographics by Neighborhood}$

This map shows 2018 data the Hispanic/Latino population in LA County by neighborhood. It is in a choropleth style. 
```{r echo=FALSE}
bins <- stats::quantile(demog.poly$hislat)
pal <- colorBin(c("#f7f7f7","#cccccc","#969696","#636363","#252525"), domain = demog.poly$hislat, bins = bins)

leaflet(demog.poly) %>%
  addProviderTiles("OpenStreetMap.Mapnik") %>%
  addPolygons(
    fillColor = ~pal(hislat),
    weight = 2,
    fillOpacity = .8,
    color ="white"
  ) %>%
  addLegend(position = "topright" , pal = pal, values = demog.poly$hislat, title = "Hispanic/Latinos by Neighborhood", labFormat = labelFormat (digits = 0))

```

$\textbf{All data found at:}$
https://geohub.lacity.org/